---
title: "Step-1 of Project Analyses - Exploration"
subtitle: "Example [INCOMPLETE TEMPLATE] - Screen Use & Sleep Quality"
author: "PSY150"
date: "`r Sys.Date()`"
output: 
    html_document:
      theme: sketchy
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

------------------------------------------------------------------------

### Data source is simulated or fabricated to reflect finding of following research study:

Jniene et al. (2019). *Perception of sleep disturbances due to bedtime use of blue light‐emitting devices and its impact on habits and sleep quality among young medical students.* 

------------------------------------------------------------------------

```{r}

library(tidyverse)
library(janitor)
library(report)
library(gt)
library(psych)
library(sjPlot)
library(corrplot)

source("load_plot_functions.R")

```

------------------------------------------------------------------------

### Read in data & prepare for analyses

------------------------------------------------------------------------

Read in data 
```{r}
raw_data <- read_csv(here("data", "blue_light_sleep_sim25.csv")) %>%
  clean_names()
```

Prepare data

```{r}

sleep_data <- raw_data %>% 
    select(
        sleep_quality = sleep_quality_psqi, 
        sleep_hrs = sleep_duration_hours, 
        fatigue_weekly = fatigue_on_waking_perweek,
        device_use = device_use_minutes_bedtime,
        sleep_efficiency = sleep_efficiency_pct,
        age, room_lights_off, sex, use_reason, chronotype) %>% 
  mutate(
    female = factor(sex, levels = c("male","female")),
    use_reason = factor(use_reason, levels = c("reading","leisure")), 
    chronotype = factor(chronotype, levels = c("morning","intermediate","evening"))
  )

```

------------------------------------------------------------------------

### Take a look at descriptive statistics using `describe()`

- Descriptive stats are useful for describing continuous or ordinal variables
- Such as: Mean, median, mode, range, skew, kurtosis...
- `describe()` will return an error if you include non-numeric variables
- The 'mean' of a binary variable will the portion of 1's for variable
- Descriptive stats do not work with nominal/categorical variables 

------------------------------------------------------------------------

```{r}
sleep_data %>%  
  select(sleep_quality, 
         sleep_hrs, 
         fatigue_weekly,
         device_use,
         age, room_lights_off) %>%
  describe()

```

------------------------------------------------------------------------

### Look at distributions of continuous variables 

- Use `hist_plot()` to look at distribution or shape of variables
- Works well with continuous or ordinal variables (e.g., Likert scales)
- You should plot distributions of ALL variables included in your analyses

------------------------------------------------------------------------

Plot distribution of `sleep_quality` with a histogram
```{r}

hist_plot(
  data = sleep_data,
  binwidth=1,
  x_var = sleep_quality,
  x_label = "sleep quality psqi",
  fill = "yellow", color = "purple",
  break_min =0, break_max =24 , break_by = 1,
  mean_line = TRUE, median_line = TRUE,
  skew_value = TRUE, kurtosis_value = TRUE,
  title   = "")

```

Plot distribution of `device_use` with a histogram
```{r}

hist_plot(
  data = sleep_data,
  binwidth=10,
  x_var = device_use,
  x_label = "",
  fill = "blue3", color = "red",
  break_min =0, break_max =170 , break_by = 10,
  mean_line = TRUE, median_line = TRUE,
  skew_value = TRUE, kurtosis_value = TRUE,
  title   = ""
)

```

Plot distribution of `sleep_efficiency` with a histogram
```{r}

hist_plot(
  data = sleep_data,
  binwidth=1,
  x_var = sleep_efficiency,
  x_label = "sleep efficiency",
  fill = "lightblue", color = "darkorange",
  break_min = TRUE, median_line = TRUE, 
  skew_value = TRUE, kurtosis_value = TRUE,
  mean_line = TRUE, 
  
)
```

Plot distribution of `fatigue_weekly` with a histogram
```{r}
hist_plot(
  data = sleep_data,
  binwidth=1,
  x_var = fatigue_weekly,
  x_label = "Self-Reported Fatigue on Waking (per week)",
  fill = "blue3", color = "red3",
  break_min = TRUE, median_line = TRUE, 
  skew_value = TRUE, kurtosis_value = TRUE,
  mean_line = TRUE)

```


------------------------------------------------------------------------

### Look at bivariate associations using scatterplots

- Scatterplots are used to look at associations between two variables
- `scatter_plot()` is best used to describe two continuous and/or ordinal variables
- NOTE: `scatter_plot()` prints Pearson's `r` correlation but not significance (`p`)

------------------------------------------------------------------------

What is the association between `device_use` and `sleep_quality`?
```{r}
scatter_plot(
  data = sleep_data,
  x_var = device_use,
  x_label = "Screentime Use Before Bed (Minutes)",
  y_var = sleep_quality,
  y_label = "Sleep Quality PSQI Score",
  point_color = "green3",
  corr_line = TRUE,
  title   = "Is screentime before bed associated with sleep quality?"
)
```

What is the association between `device_use` and `fatigue_weekly`?
```{r}
scatter_plot(
  data = sleep_data,
  x_var = device_use,
  x_label = "Minutes of Screentime Before Bed",
  y_var = fatigue_weekly,
  y_label = "Self-Reported Fatigue on Waking (per week) ",
  point_color = "red3",
  corr_line = TRUE,
  title   = "Is screentime before bed associated with sleep quality?"
)
```

------------------------------------------------------------------------

Explore multiple associations using correlation tables & plots

------------------------------------------------------------------------

Make an APA formatted correlation table 
```{r}

corr_vars <- sleep_data %>% 
  select(sleep_quality, device_use, fatigue_weekly,
         sleep_efficiency,
         sleep_hrs)

corr_vars %>% 
tab_corr(
  triangle = "lower",
  digits = 2
)
```

Make a correlation plot 
```{r}
corr_matrix <- cor(corr_vars, use = "pairwise.complete.obs",
                   method = "pearson")

corrplot(
  corr_matrix, 
  method = "color",   
  type = "upper", diag = FALSE,       
  tl.col = "black", tl.srt = 35)
```

------------------------------------------------------------------------

### Exploring Categorical Variables 

- Look at frequencies or counts of categorical variables using `tabyl()`
- In `tabyl()` the `n` is the count or frequency 
- `percent` tells you the percent of sample for each level of factor

------------------------------------------------------------------------

How many respondents reported `sex` as `male` and `female` in sample?
```{r}
tabyl(sleep_data$female)
```

What are the counts of early, intermediate, and evening risers (`chronotype`)?
```{r}
tabyl(sleep_data$chronotype)
```

------------------------------------------------------------------------

### Plotting counts of a categorical variable with 2 or more levels using `bar_plot()`

------------------------------------------------------------------------

Plot relative counts of `early`, `intermediate`, and `evening` riser's in the sample
```{r}

bar_plot(
  data = sleep_data,
  x_var = chronotype,
  x_label = "Self-Reported Preferred Wake Up Time (Chronotype)",
  title = "How many morning people, intermediate riser's, and late riser's are in sample?")

```

------------------------------------------------------------------------

### Looking at associations between a continuous var (Y) and categorical predictor (X)

`mean_diff_barplot()` arguments (inputs): 
- `data`: The data object to use for plotting
- `outcome`: The continuous variable you want to plot on the `Y-axis`
- `group_var`: The categorical factor you want to plot on the `X-axis`
- `colors`: The number of colors must match number of factor levels

------------------------------------------------------------------------

What is the difference in average sleep quality by `use_reason` (`reading` vs `leisure`)?
```{r}

mean_diff_barplot(
  data = sleep_data, 
  outcome = "sleep_quality",
  group_var = "use_reason",
  colors = c("cyan", "blue"),
  title = "Does sleep quality differ if you use phone for reading vs. leisure?"
)

```

What is the difference in average sleep quality by `chronotype` group?

```{r}

mean_diff_barplot(
  data = sleep_data, 
  outcome = "sleep_quality",
  group_var = "chronotype",
  colors = c("purple","aquamarine","darkblue"),
  title = "Does sleep quality differ if you use phone for reading vs. leisure?"
)

```


