---
title: "Lab3: Investigate Distributions, Variable Associations, & Correlations"
subtitle: "PSY 150: Lab Exercise "
author: "ADD_YOUR_NAME_HERE"
date: "`r Sys.Date()`"
output: 
    html_document:
      theme: sketchy
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval=TRUE)
```

------------------------------------------------------------------------

### Load packages

------------------------------------------------------------------------

```{r}

library(psych)
library(tidyverse)
library(janitor)
library(ggpubr)
library(ggrepel)
library(here)
library(palmerpenguins)
library(corrplot)
library(sjPlot)

```


------------------------------------------------------------------------

### Load Plot Functions 

------------------------------------------------------------------------

```{r}
source("load_plot_functions.R")
```

------------------------------------------------------------------------

### Keyboard shortcuts (Use them!)

> MAC / *WINDOWS*

- `Run code` (at cursor): COMMAND/*CONTROL* + RETURN/*ENTER*
- `Pipe operator` (`%>%`): COMMAND/*CONTROL* + SHIFT + M
- `Add code chunk`: COMMAND/*CONTROL* + OPTION/*ALT* + I
- `Assignment operator` (`<-`): OPTION/*ALT* + MINUS("-") 
- `Search`: COMMAND/*CONTROL* + F

------------------------------------------------------------------------

### Penguins Data Source: {`palmerpenguins`}

The Palmer Penguins dataset contains real measurements of penguins observed on three islands in the Palmer Archipelago, Antarctica (Horst et al., 2020).

**Source:** https://allisonhorst.github.io/palmerpenguins/

------------------------------------------------------------------------

Load Palmer Penguin data set
```{r}
penguin_data <- palmerpenguins::penguins
```

------------------------------------------------------------------------

### What is the bimodal distribution `flipper_length_mm` hiding? 

------------------------------------------------------------------------

```{r}
penguin_data %>% 
ggplot(aes(x = flipper_length_mm, fill = species)) +
  geom_histogram(
    alpha = 0.7, 
    position = "identity") +
  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  labs(x = "Flipper length (mm)",
       y = "Frequency",
       title = "Penguin flipper lengths") + 
  theme_classic_grid()
```

------------------------------------------------------------------------

### Data source:

> In this part of the lab we are using data that was simulated in R to follow trends found in behavioral science data. The data is not real survey data so do not take it too seriously! 

------------------------------------------------------------------------

### Read in the `CSV` spreadsheet named `sim_college_data.csv` located in the `data` folder

```{r}

college_data <- read_csv(here("data", "sim_college_data.csv"))

```

Note: Always take a look at the data!

------------------------------------------------------------------------

### Take a look at the variables available in `college_data`

------------------------------------------------------------------------

```{r}
names(college_data)
```

------------------------------------------------------------------------

### Investigating frequency distributions using histogram plots

How many minutes does it take to park at SBCC?

------------------------------------------------------------------------

```{r}

hist_plot(
  data = college_data,
  x_var = park_min,
  x_label = "Minutes to Find Parking at SBCC",
  binwidth=2,
  fill = "yellow",
  color = "black",
  break_min = 0, break_max = 35, break_by = 10,
  mean_line = TRUE, median_line = TRUE,
  skew_value = TRUE, kurtosis_value = TRUE,
  title = "How many minutes does it take to park at SBCC?",
  )  

```

**Question 1:** Describe the distribution: 

- Is the variable skewed & if so which direction is it skewed? 
- How does the mean compare to the median, and what explains this difference?

**Response:** __________________________________________

------------------------------------------------------------------------

### What is the caffiene intake for students at SBCC per day?

------------------------------------------------------------------------

```{r}

hist_plot(
  data = college_data,
  x_var = coff_cups,
  x_label = "Cups of caffiene Per Day",
  binwidth=1,
  fill = "sienna3", color = "black",
  break_min = 0, break_max = 7, break_by = 1,
  mean_line = TRUE, median_line = TRUE,
  skew_value = TRUE, kurtosis_value = TRUE,
  title   = "How many caffienated drinks do students drink per day?"
)


```

**Question 2:** What is the **mode** for the variable `caffeinated drinks`? 

**Response:** __________________________________________

------------------------------------------------------------------------

### Are you satisfied with the location of the SBCC campus?

------------------------------------------------------------------------

```{r}

hist_plot(
  data = college_data,
  binwidth=1,
  x_var = view_satisfy,
  x_label = "Satisfaction Rating (1-10)",
  fill = "green3", color = "cyan2",
  break_min = 0, break_max = 10, break_by = 1,
  mean_line = TRUE, median_line = TRUE,
  skew_value = TRUE, kurtosis_value = TRUE,
  title   = "Are you satisfied with the location of the SBCC campus?"
)

```

**Question 3:** Is the variable skewed & if so which direction is it skewed? 

**Response:** __________________________________________

------------------------------------------------------------------------

### Pick a variable to look at it's distribution?

------------------------------------------------------------------------

```{r, eval=FALSE}

hist_plot(
  data = college_data,
  binwidth=***,
  x_var = ***,
  x_label = "***",
  fill = "***", color = "***",
  #break_min = , break_max = , break_by = ,
  mean_line = TRUE, median_line = TRUE,
  skew_value = TRUE, kurtosis_value = TRUE,
  title   = ""
)

```

**Question 4:** Describe the variables shape:

**Response:** __________________________________________


------------------------------------------------------------------------

### Two variable associations & correlation

Does coffee help you stay awake?

------------------------------------------------------------------------

```{r}

scatter_plot(
  data = college_data,
  x_var = coff_cups,
  x_label = "Cups of Coffee (per day)",
  y_label = "Awake Hours (per day)",
  y_var = awake_hrs,
  point_color = "sienna3",
  corr_line = TRUE,
  title   = "Coffee vs. Awake Hours",
)

```


**Question 5:** Describe the correlation:

- How strong is the correlation (weak, moderate, strong)? 
- Is the correlation positive or negative?

**Response:** __________________________________________

------------------------------------------------------------------------

### Are the number of laptop stickers associated with exam score?

------------------------------------------------------------------------

```{r}

scatter_plot(
  data = college_data,
  x_var = laptop_stickers,
  x_label = "Number of Laptop Stickers",
  y_var = exam_score,
  y_label = "Exam Score (%)",
  point_color = "dodgerblue3",
  corr_line = TRUE,
  title   = "Laptop Stickers vs. Exam Score"
)

```

**Question 6:** Describe the correlation (strength/direction). Is this association meaningful? 

**Response:** __________________________________________


------------------------------------------------------------------------

### Are the number of texts you send related to the number you receive?

------------------------------------------------------------------------

```{r}

scatter_plot(
  data = college_data,
  x_var = text_sent,
  x_label = "Texts Sent (per day)",
  y_var = text_recv,
  y_label = "Texts Received (per day)",
  point_color = "darkgreen",
  corr_line = TRUE,
  title   = "Texts Sent vs. Received"
)

```

**Question 7:** How strong is the correlation (weak, moderate, strong)? 

**Response:** __________________________________________

------------------------------------------------------------------------

### Choose 2 variables and explore how they are related

------------------------------------------------------------------------

```{r, eval=FALSE}

scatter_plot(
  data = college_data,
  x_var = ***,
  x_label = "***",
  y_var = ***,
  y_label = "***",
  point_color = "***",
  corr_line = TRUE,
  title   = ""
)

```

**Question 8:** Describe the pattern you see:

**Response:** __________________________________________


------------------------------------------------------------------------

### Create a correlation plot

------------------------------------------------------------------------

Estimate correlations & significance
```{r}

# Select variables to plot 
sample_corr <- corr.test(
  college_data %>% 
  select(react_time:netfilx_hrs,   
         study_hrs:text_recv,
         hw_hours,beach_hrs,sand_car),
  adjust = "none" 
)

```

Create a correlation plot with significance 
```{r}

corrplot(
  sample_corr$r, 
  p.mat = sample_corr$p, 
  method = "color",   
  type = "upper", diag = FALSE,       
  insig = "label_sig", pch.cex  = 0.8,
  sig.level = c(.001, .01, .05),
  tl.col = "black", tl.srt = 35)

```

**Question 9:** Select a pair of variables, describe the strength of their correlation, and state if the p-value indicates significance:

**Response:** __________________________________________

------------------------------------------------------------------------

### ### Create a correlation table

------------------------------------------------------------------------

```{r}

college_data %>% 
  select(coff_cups, digit_recall,exam_score,
         study_hrs,sleep_hrs, netfilx_hrs) %>% 
tab_corr(
  triangle = "lower",
  digits = 2
)
```

**Question 10:** Select a different pair of variables, describe the strength of their correlation, and state if the p-value indicates significance.

**Response:** __________________________________________


