---
title: "Lab 2: Exploring the General Social Survey (GSS)"
subtitle: "PSY 150: Data Wrangling + Differences in Means & Variation"
author: "ADD_YOUR_NAME_HERE"
date: "`r Sys.Date()`"
output: 
    html_document:
      theme: sketchy
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

------------------------------------------------------------------------

### Lab goals

By the end of this lab you will be able to:

* Load an applied social‑science dataset (`GSS`) from an R package
* View rows/columns and simple summaries
* Rename columns to clearer names
* Select a subset of columns you care about
* Filter rows to focus on specific groups
* Convert variables to factors (categories)
* Label and re‑order factor (category) levels
* Create plots to look at differences in means & variation across groups

------------------------------------------------------------------------

### Data source:

In this lab we use an open source data-set called **General Social Survey (GSS)**. It contains responses from U.S. adults from 2000–2014 on topics like age, marital status, political party, religion, income, and hours of TV watched. These are real survey responses.

------------------------------------------------------------------------

### 📦 Packages: Install once... load libraries every time!

-   If a `yellow banner` appears at top of screen, asking you to install packages click `YES`
-   Packages include functions other R users have made for us to make our lives easier!

------------------------------------------------------------------------

```{r}
# Step 1: Load packages

library(tidyverse)
library(janitor)
library(psych)
library(gt)
library(gtsummary)
library(ggpubr)
library(naniar)
library(ggthemes)

```

### STEP 0: `Only` do this if a package(s) above did not load (turns red)

```{r}
# DIRECTIONS:
# 1. Each line starts with a "#" symbol, which means it is ignored by R (a comment).
# 2. To "UNCOMMENT” a line, delete the "#" at the beginning of that line.
# 3. CLICK RUN: This will install the package with the matching name in quotes.
# 4. "RE-COMMENT":After installation is finished, put the "#" back at the start of the line. 
#    *This makes the text green again so it won’t try to install every time.*
# 5. RETURN TO PREVIOUS CODE BLOCK and RE-RUN  again to load the packages.

# install.packages("tidyverse")
# install.packages("janitor")
# install.packages("psych")
# install.packages("gt")
# install.packages("gtsummary")
# install.packages("ggpubr")
# install.packages("naniar")
# install.packages("ggthemes")

# NOTE: AFTER install finishes you must return to code block above & RE-RUN
```



------------------------------------------------------------------------

### Keyboard shortcuts (Use them!)

> MAC / *WINDOWS*

- `Run code` (at cursor): COMMAND/*CONTROL* + RETURN/*ENTER*
- `Pipe operator` (`%>%`): COMMAND/*CONTROL* + SHIFT + M
- `Add code chunk`: COMMAND/*CONTROL* + OPTION/*ALT* + I
- `Assignment operator` (`<-`): OPTION/*ALT* + MINUS("-") 
- `Search`: COMMAND/*CONTROL* + F

------------------------------------------------------------------------

### Read in the data

------------------------------------------------------------------------

Load in the data set called `gss_cat` from the `forcats` package
```{r}
gss_raw <- gss_cat
```

Note: Always take a look at the data!

------------------------------------------------------------------------

### Prepare data for exploring

------------------------------------------------------------------------

```{r}

names(gss_raw)

```

**Qn:** What are some of the variables available in the data `gss_raw`?

**Response:**__________________________________ 


```{r}

gss_vars <- gss_raw %>% 
  select(year, age, marital, relig, tvhours) %>% 
  rename(tv_hours = tvhours) %>% 
  drop_na(tv_hours)

```

**Qn:** What does the function `select()` do?

**Response:**__________________________________


**Qn:** What does `NA` stand for and what does `drop_na(tv_hours)` do?

**Response:**__________________________________


```{r}

glimpse(gss_vars)

```

**Qn:** What does the glimpse output tell us about `gss_vars` now?

**Response:**__________________________________

------------------------------------------------------------------------

### Age distribution: How old are the GSS survey participants?

------------------------------------------------------------------------

Take a look at the age range in the GSS survey
```{r, eval=FALSE}

tabyl(gss_vars$age)

```


```{r}

gss_youth <- gss_vars %>% 
  select(year, age, tv_hours) %>% 
  filter(age <= 20)

```

**Qn:** What does the `filter()` function do?

**Response:**__________________________________

**Qn:** What ages are included in the new data.frame we created `gss_youth`, and roughly how many observations does it have?

**Response:**__________________________________

------------------------------------------------------------------------

### How many survey respondents do we have for each age (18-20 years)?

------------------------------------------------------------------------

```{r}
gss_youth %>%
  ggplot(aes(x = factor(age), fill = factor(age))) +
  geom_bar(color = "white") +
  labs(
    title = "How many survey respondents do we have for each age (18-20 years)?",
    x = "Age", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

------------------------------------------------------------------------

### Faceted plots: Separating plot by year surveyed from 2010 - 2014

This let's us zoom-in on the data and see how response rates vary by `year` & `age`.

------------------------------------------------------------------------

```{r}
gss_youth %>% 
filter(year <= 2010) %>% 
  ggplot(aes(x = factor(age), fill = factor(age))) +
  geom_bar(color = "white") +
  facet_wrap(~ year, ncol = 3) + # Facet plot by year
  labs(
    title = "Frequency Counts by Age - Faceted by Year",
    x = "Age", y = "Count") +
  theme(legend.position = "none")

```

**Qn:** Across the 3 years surveyed, do we see differences in counts: Which year had the highest number of participants? Which age had the least number of participants?

**Response:**__________________________________

------------------------------------------------------------------------

### TV watching habits 

------------------------------------------------------------------------

Let's look at the variable Hours of TV watched in a day.
```{r}

gss_youth %>% 
  filter(age == 18) %>% 
  select(tv_hours) %>% 
  describe() %>%  
  select(n, mean, median, sd, min, max, range)

```

**Qn:** How many hours of TV on average does this subset of the sample watch (age=18)? What is the median?

**Response:**__________________________________

------------------------------------------------------------------------

### Take a look at the distribution for TV watching 

------------------------------------------------------------------------

Create a `density curve plot` to see the shape of the `tv_hours` distribution for respondents age 18 only. 
```{r}

gss_youth %>% 
  filter(age == 18) %>% 
  ggplot(aes(x = tv_hours)) +
    geom_density(fill = "lightblue", alpha = 0.4) +
    geom_vline(xintercept = mean(gss_youth$tv_hours), color = "red", size = 1) +
    labs(title = "Density of TV Hours (Age 18)",
         x = "TV Hours per Day", y = "Density") +
  theme_minimal()

```

**Note:** This plot is called a density curve (a smoothed version of a histogram). The red line marks the mean TV hours for 18 year-old respondent subset.

**Qn:** What is the shape of the distribution of TV hours for 18-year-old respondents?

**Response:**__________________________________

------------------------------------------------------------------------

### How much TV do 18 year old's watch on average from 2000-2014?

------------------------------------------------------------------------

What is the average hiding?
```{r}
gss_youth %>%
  filter( age == 18) %>%
  summarise(mean_tv = mean(tv_hours), sd_tv = sd(tv_hours), .groups = "drop") %>%
  ggplot(aes(x = "18", y = mean_tv)) +
  geom_col(width = 0.7, fill = "white", color = "blue") +
   geom_jitter(data = gss_youth %>% filter(age == 18, !is.na(tv_hours)),
     aes(x = "18", y = tv_hours), height = 0,
     width = 0.2, alpha = 0.4, size = 2.5, color = "blue") +
  expand_limits(y = 0) +   
  coord_cartesian(ylim = c(0, 13)) + 
  labs(title = "Average Hours of TV Watched per Day",
       x = "Age", y = "Average TV Hours") +
  theme_classic()

```

**Qn:** What is the average value hiding about the variable `tv_hours`? Why could focusing only on this average be misleading?

**Response:**__________________________________

------------------------------------------------------------------------

### Grouping by age: Comparing TV watching across age (18 - 20)

------------------------------------------------------------------------

For a specific snapshot, let's focus on the year 2014 and compare the TV watching of respondents 18 - 20
```{r}
tvage_stats <- gss_youth %>%
  filter(year == 2014) %>% 
  group_by(age) %>%
  summarise(mean_tv = mean(tv_hours),
            sd_tv   = sd(tv_hours),
            .groups = "drop")
```

**Qn:** What does `group_by(age)` do when combined with `summarise()`?

**Response:**__________________________________

------------------------------------------------------------------------

Look at means for `tv_hours` grouped by age using a `barplot`

------------------------------------------------------------------------

```{r}
tvage_stats %>% 
  ggplot(aes(x = factor(age), y = mean_tv, fill = factor(age))) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(
    label = paste0("Mean = ", round(mean_tv, 1), " hours"),
            y = mean_tv + 0.1), size = 3.5) + 
  labs(title = "Mean Hours of TV Watched per Day by Age (18–24)",
       x = "Age",
       y = "Average TV Hours") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```


**Qn:** What could explain why the average TV hours differ more between 19- and 20-year-olds compared to the other age groups?

**Response:**__________________________________

------------------------------------------------------------------------

### Plotting variation: What is the mean hiding?

------------------------------------------------------------------------

```{r}
tvage_stats %>% 
ggplot(aes(x = factor(age), y = mean_tv,
                      color = factor(age))) +
  geom_col(width = 0.7, show.legend = FALSE, fill = "white") +
  geom_jitter(
    data = gss_youth %>% filter(!is.na(tv_hours)),
    aes(x = factor(age), y = tv_hours, color = factor(age)),
    width = 0.15, height = 0,
    alpha = 0.4, size = 1.7) +
  geom_errorbar(
    aes(ymin = pmax(0, mean_tv - sd_tv), ymax = mean_tv + sd_tv),
    width = 0.15,
    linewidth = 0.3,
    color = "black") +
  labs(
    title = "Mean TV Hours per Day by Age (18–24)",
    x = "Age", y = "Average TV Hours"
  ) +
  theme_minimal() + 
  guides(color = "none")

```


**Qn:** From the plot, describe the difference is variation for TV watching, across the three groups. Which groups mean is most influenced by outliers (extreme values)?

**Response:**__________________________________

------------------------------------------------------------------------

### Do people who are religious watch more (or less) TV?

------------------------------------------------------------------------

Prepare data by creating a new variable `religion2` with two comparison groups. 
```{r}

# tabyl(gss_vars$relig)

gss_tv_relig <- gss_vars %>% 
  filter(year == 2014) %>%
  mutate(
    religion2 = case_when(
      relig == "None" ~ "no_religion",
      relig %in% c("No answer", "Don't know", "Not applicable") ~ NA_character_,
      TRUE ~ "religion"),
    religion2 = factor(religion2, 
           levels = c("no_religion", "religion"),
           labels = c("No Religion", "Religion"))) %>% 
  drop_na(religion2)
```

**Qn:** How did we create the variable `religion2` with two categories (factor levels)?

**Response:**__________________________________


Compare TV hours between these two groups with a summary table
```{r}
gss_tv_relig %>% 
  select(religion2,tv_hours) %>% 
  tbl_summary(
    by =religion2,
    statistic = list(all_continuous() ~ "{mean} ({sd})"),
    digits = all_continuous() ~ 1)
  
```


------------------------------------------------------------------------

### Visualizing TV watching by religion

------------------------------------------------------------------------

Step 1: Calculate group-level statistics (means and SDs)
```{r}
relig_stats <- gss_tv_relig %>%
  group_by(religion2) %>%
  summarise(mean_tv = mean(tv_hours), sd_tv = sd(tv_hours), .groups = "drop")
```

Step 2: Plot the averages, standard deviations, and raw data points
```{r}
relig_stats %>% 
ggplot(aes(x = religion2, y = mean_tv, color = religion2)) +
  geom_col(width = 0.6, fill = "white") +
  geom_jitter(data = gss_tv_relig,
    aes(x = religion2, y = tv_hours, color = religion2),
    width = 0.25, height = 0.1, alpha = 0.35, size = 1.6) +
  labs(
    title = "Average TV Hours per Day by Religious Belief (2014)",
    x = "Religion Group", y = "Mean TV Hours") +
  geom_errorbar(aes(ymin = mean_tv - sd_tv,
                    ymax = mean_tv + sd_tv),
                width = 0.15, linewidth = 0.3, color = "black") +
  coord_cartesian(ylim = c(0, NA)) +
  theme_minimal() +
  guides(color = "none", fill = "none")

```

**Qn:** Based on the plot, how do the average TV hours and the amount of variation compare between people with No Religion and those with a Religion?

**Response:**__________________________________


